\documentclass[natbib]{sigplanconf}

\usepackage[usenames]{color}
\definecolor{citationcolour}{rgb}{0,0.4,0.2}
\definecolor{linkcolour}{rgb}{0,0,0.8}
\usepackage{hyperref}
\hypersetup{colorlinks=true,
            urlcolor=linkcolour,
            linkcolor=linkcolour,
            citecolor=citationcolour,
            pdftitle=Productive Coprogramming with Guarded Recursion,
            pdfauthor={Robert Atkey, Conor McBride},
            pdfkeywords={}}  
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}

\title{Productive Coprogramming with Guarded Recursion}

\authorinfo{Robert Atkey}
           { }
           {bob.atkey@gmail.com}

\authorinfo{Conor McBride}
           {University of Strathclyde}
           {Conor.McBride@strath.ac.uk}

\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{stmaryrd}

% Colouring in...
\definecolor{highlight}{rgb}{0.8,0.1,0.1}
\definecolor{cons}{rgb}{0.8,0.1,0.2}
\definecolor{elim}{rgb}{0.1,0.5,0.35}
\definecolor{defnd}{rgb}{0.1,0.2,0.8}
\definecolor{greybg}{rgb}{0.8,0.8,0.8}
\newcommand{\highlight}[1]{\textcolor{highlight}{#1}}
\newcommand{\cons}[1]{\textcolor{cons}{\textsf{#1}}}
\newcommand{\elim}[1]{\textcolor{elim}{\textsf{#1}}}
\newcommand{\defnd}[1]{\mathit{#1}}
\newcommand{\kw}[1]{\textbf{#1}}
\newcommand{\tyname}[1]{\textrm{#1}}
\newcommand{\ident}[1]{\textit{#1}}
\newcommand{\defn}[1]{\textcolor{defnd}{\ident{#1}}}

% other stuff
\newcommand{\sepbar}{\mathrel|}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\op}{\mathsf{op}}
\newcommand{\sortType}{\texttt{\textup{type}}}

\newcommand{\delay}[1]{\mathord{\rhd\kern-0.4em^{#1}}}
\newcommand{\delayX}[1]{\mathord{\rhd\kern-0.1em^{#1}}}

\newcommand{\clkSem}[1]{\llbracket #1 \rrbracket}
\newcommand{\ClkPER}{\mathrm{ClkPER}}
\newcommand{\PER}{\mathrm{PER}}
\newcommand{\semCons}[1]{\mathsf{#1}}
\newcommand{\tySem}[1]{\llbracket #1 \rrbracket}

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}

\newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
\newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}

\newcommand{\etal}{\textit{et} \textit{al}.}

\begin{document}

\exclusivelicense
\conferenceinfo{ICFP~'13}{September 25--27, 2013, Boston, MA, USA}
\copyrightyear{2013}
\copyrightdata{978-1-4503-2326-0/13/09}
\doi{2500365.2500597}

\maketitle

\begin{abstract}
  Total functional programming offers the beguiling vision that, just
  by virtue of the compiler accepting a program, we are guaranteed
  that it will always terminate. In the case of programs that are not
  intended to terminate, e.g., servers, we are guaranteed that
  programs will always be \emph{productive}. Productivity means that,
  even if a program generates an infinite amount of data, each piece
  will be generated in finite time. The theoretical underpinning for
  productive programming with infinite output is provided by the
  category theoretic notion of final coalgebras. Hence, we speak of
  \emph{co}programming with non-well-founded \emph{co}data, as a dual
  to programming with well-founded data like finite lists and trees.

  Systems that offer facilities for productive coprogramming, such as
  the proof assistants Coq and Agda, currently do so through syntactic
  guardedness checkers, which ensure that all self-recursive calls are
  guarded by a use of a constructor. Such a check ensures
  productivity. Unfortunately, these syntactic checks are not
  compositional, and severely complicate coprogramming.

  Guarded recursion, originally due to Nakano, is tantalising as a
  basis for a flexible and compositional type-based approach to
  coprogramming. However, as we show, guarded recursion by itself is
  not suitable for coprogramming due to the fact that there is no way
  to make finite observations on pieces of infinite data. In this
  paper, we introduce the concept of \emph{clock variables} that index
  Nakano's guarded recursion. Clock variables allow us to ``close
  over'' the generation of infinite codata, and to make finite
  observations, something that is not possible with guarded recursion
  alone.
\end{abstract}

\category{D.1.1}{Programming techniques}{Applicative (functional)
  programming} \category{D.2.4}{Software Engineering}{Software/Program
  Verification} \category{D.3.3}{Programming Languages}{Language
  Constructs and Features---Data types and structures}

\terms
  Languages, Theory, Types, Recursion

\keywords
  coalgebras, corecursion, guarded recursion, total functional programming

\input{introduction.tex}
\input{type-system.tex}
\input{semantics.tex}
\input{fixpoint-types.tex}
\input{conclusions.tex}

\paragraph{Acknowledgements} We would like to thank Lars Birkedal,
Rasmus M{\o}gelberg, and Paula Severi for extremely useful comments
and discussions.

\begin{thebibliography}{24}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi: #1}\else
  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

\bibitem[Abel(2004)]{abel04termination}
A.~Abel.
\newblock Termination checking with types.
\newblock \emph{ITA}, 38\penalty0 (4):\penalty0 277--319, 2004.

\bibitem[Appel and McAllester(2001)]{appel01indexed}
A.~W. Appel and D.~A. McAllester.
\newblock An indexed model of recursive types for foundational proof-carrying
  code.
\newblock \emph{ACM Trans. Program. Lang. Syst.}, 23\penalty0 (5):\penalty0
  657--683, 2001.

\bibitem[Bird(1984)]{bird84circular}
R.~S. Bird.
\newblock {Using Circular Programs to Eliminate Multiple Traversals of Data}.
\newblock \emph{Acta Informatica}, 21:\penalty0 239--250, 1984.

\bibitem[Birkedal and M{\o}gelberg(2013)]{birkedal13intensional}
L.~Birkedal and R.~E. M{\o}gelberg.
\newblock Intensional type theory with guarded recursive types qua fixed points
  on universes.
\newblock In \emph{ACM/IEEE Symposium on Logic in Computer Science (LICS
  2013)}, 2013.

\bibitem[Birkedal et~al.(2010)Birkedal, Schwinghammer, and
  St{\o}vring]{birkedal10metric}
L.~Birkedal, J.~Schwinghammer, and K.~St{\o}vring.
\newblock A metric model of guarded recursion.
\newblock In \emph{Presented at FICS 2010}, 2010.

\bibitem[Birkedal et~al.(2012)Birkedal, M{\o}gelberg, Schwinghammer, and
  St{\o}vring]{birkedal12first}
L.~Birkedal, R.~E. M{\o}gelberg, J.~Schwinghammer, and K.~St{\o}vring.
\newblock First steps in synthetic guarded domain theory: step-indexing in the
  topos of trees.
\newblock \emph{Log. Meth. in Computer Science}, 8\penalty0 (4), 2012.

\bibitem[Capretta(2005)]{capretta05general}
V.~Capretta.
\newblock General recursion via coinductive types.
\newblock \emph{Log. Meth. in Computer Science}, 1\penalty0 (2), 2005.

\bibitem[Danielsson(2010)]{danielsson10beating}
N.~A. Danielsson.
\newblock Beating the productivity checker using embedded languages.
\newblock In \emph{Workshop on Partiality and Recursion in Interactive Theorem
  Provers (PAR 2010)}, volume~43 of \emph{EPTCS}, pages 29--48, 2010.

\bibitem[Danielsson and Altenkirch(2009)]{danielsson09mixing}
N.~A. Danielsson and T.~Altenkirch.
\newblock Mixing induction and coinduction.
\newblock Draft, 2009.

\bibitem[Dreyer et~al.(2011)Dreyer, Ahmed, and Birkedal]{dreyer11logical}
D.~Dreyer, A.~Ahmed, and L.~Birkedal.
\newblock Logical step-indexed logical relations.
\newblock \emph{Log. Meth. in Computer Science}, 7\penalty0 (2), 2011.

\bibitem[Ghani et~al.(2009)Ghani, Hancock, and
  Pattinson]{ghani09representations}
N.~Ghani, P.~Hancock, and D.~Pattinson.
\newblock Representations of stream processors using nested fixed points.
\newblock \emph{Log. Meth. in Computer Science}, 5\penalty0 (3), 2009.

\bibitem[Gim{\'e}nez(1994)]{gimenez94codifying}
E.~Gim{\'e}nez.
\newblock Codifying guarded definitions with recursive schemes.
\newblock In \emph{Types for Proofs and Programs, International Workshop
  TYPES'94}, volume 996 of \emph{Lecture Notes in Computer Science}, pages
  39--59, 1994.

\bibitem[Hutton and Jaskelioff(2011)]{hutton11representing}
G.~Hutton and M.~Jaskelioff.
\newblock Representing contractive functions on streams.
\newblock Submitted, 2011.

\bibitem[Krishnaswami and Benton(2011{\natexlab{a}})]{krishnaswami11semantic}
N.~R. Krishnaswami and N.~Benton.
\newblock A semantic model for graphical user interfaces.
\newblock In \emph{ACM SIGPLAN international conference on Functional
  Programming, ICFP 2011}, pages 45--57, 2011{\natexlab{a}}.

\bibitem[Krishnaswami and
  Benton(2011{\natexlab{b}})]{krishnaswami11ultrametric}
N.~R. Krishnaswami and N.~Benton.
\newblock Ultrametric semantics of reactive programs.
\newblock In \emph{IEEE Symposium on Logic in Computer Science, LICS 2011},
  pages 257--266, 2011{\natexlab{b}}.

\bibitem[Launchbury and {Peyton Jones}(1994)]{launchbury94lazy}
J.~Launchbury and S.~L. {Peyton Jones}.
\newblock Lazy functional state threads.
\newblock In \emph{Proceedings of the ACM SIGPLAN'94 Conference on Programming
  Language Design and Implementation (PLDI)}, pages 24--35, 1994.

\bibitem[Loader(1997)]{loader97equational}
R.~Loader.
\newblock {Equational Theories for Inductive Types}.
\newblock \emph{Annals of Pure and Applied Logic}, 84\penalty0 (2):\penalty0
  175--217, 1997.

\bibitem[McBride and Paterson(2008)]{mcbride08applicative}
C.~McBride and R.~Paterson.
\newblock Applicative programming with effects.
\newblock \emph{J. Funct. Prog.}, 18\penalty0 (1):\penalty0 1--13, 2008.

\bibitem[Moggi and Sabry(2001)]{DBLP:journals/jfp/MoggiS01}
E.~Moggi and A.~Sabry.
\newblock Monadic encapsulation of effects: a revised approach (extended
  version).
\newblock \emph{J. Funct. Prog.}, 11\penalty0 (6):\penalty0 591--627, 2001.

\bibitem[Nakano(2000)]{nakano00modality}
H.~Nakano.
\newblock A modality for recursion.
\newblock In \emph{IEEE Symposium on Logic in Computer Science (LICS 2000)},
  pages 255--266, 2000.

\bibitem[Pitts(1994)]{PittsAM:compavm}
A.~M. Pitts.
\newblock Computational adequacy via `mixed' inductive definitions.
\newblock In \emph{Mathematical Foundations of Programming Semantics, Proc.\
  9th Int.\ Conf.}, volume 802 of \emph{Lecture Notes in Computer Science},
  pages 72--82. Springer-Verlag, Berlin, 1994.

\bibitem[Severi and {de Vries}(2012)]{severi12pure}
P.~Severi and F.-J. {de Vries}.
\newblock Pure type systems with corecursion on streams: from finite to
  infinitary normalisation.
\newblock In \emph{ACM SIGPLAN International Conference on Functional
  Programming, ICFP'12}, 2012.

\bibitem[Smyth and Plotkin(1982)]{smyth82category}
M.~B. Smyth and G.~D. Plotkin.
\newblock The category-theoretic solution of recursive domain equations.
\newblock \emph{SIAM J. Comput.}, 11\penalty0 (4):\penalty0 761--783, 1982.

\bibitem[Tarski(1955)]{tarski55lattice}
A.~Tarski.
\newblock A lattice-theoretical fixpoint theorem and its applications.
\newblock \emph{Pacific Journal of Mathematics}, 5\penalty0 (2):\penalty0
  285--309, 1955.

\end{thebibliography}


% \bibliographystyle{plainnat}
% \bibliography{paper}

%\appendix
%\input{appendix.tex}


\end{document}
